Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-08-23 analysis/paper/paper.Rmd
Added references and smaller adjustments in text

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01]. 
The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå. 
All Raman spectra were baseline corrected using modified polynomial fitting in R (R Core Team, 2017) using the ChemoSpec package [@hanson16]. The spectra were then mean-centred and normalized using Standard Normal Variate (Shaver, 2001: 284-285) in the Prediktera Evince software.

@@

## NIR - near infrared range
No distinct peaks occurred in the 600 – 1 000 nm region, and thus the analysis focused on the 1 000 – 2 500 nm region to lessen the impact of noise. The PCA demonstrates a similar pattern to the one observed in the visual region (fig. \@ref(fig:nir-pca)). Again the first two components have an eigenvalue above 1 with cumulative explained variance of ~99.9 %. While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups. The upper left group houses the main bulk of the *brecciated quartz*, as well as the almost translucent and opaque material with a beige tint. The bottom left group instead features the majority of the white *quartz-coloured* material. In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm. In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO3 [@clark_etal90]. 
The average spectra for the three groups show that the *dark* group has the highest absorbance with less distinct peaks, although they are noticeable. The absorbance in the *brecciated*/*translucent* group is slightly higher in the lower band regions than compared to the *white* group, but with decreased absorbance in the higher regions. While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.

@@

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. n represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}

>>>>

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01]. 
The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå. 
All Raman spectra were baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16]. The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] in the Prediktera Evince software.

@@

## NIR - near infrared range
No distinct peaks occurred in the 600 – 1 000 nm region, and thus the analysis focused on the 1 000 – 2 500 nm region to lessen the impact of noise. The PCA demonstrates a similar pattern to the one observed in the visual region (fig. \@ref(fig:nir-pca)). Again the first two components have an eigenvalue above 1 with cumulative explained variance of ~99.9 %. While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups. The upper left group houses the main bulk of the *brecciated quartz*, as well as the almost translucent and opaque material with a beige tint. The bottom left group instead features the majority of the white *quartz-coloured* material. In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm. In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO^3^ [@clark_etal90]. 
The average spectra for the three groups show that the *dark* group has the highest absorbance with less distinct peaks, although they are noticeable. The absorbance in the *brecciated*/*translucent* group is slightly higher in the lower band regions than compared to the *white* group, but with decreased absorbance in the higher regions. While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.

@@

```{r raman-average, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* around 1 300 cm-1 (D band) and 1 600 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/raman_average_spectra_dark_20210823.png')
```

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
